XXX: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools

### 2023-11-28 paper/PaperCEUS/PaperCEUS.tex
paper revs from FM
@filmoura

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={biclaR: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}

@@

  \title{biclaR: Estimating the socio-environmental impacts of car
substitution by bicycle and public transit using open tools}
    \author[CERIS]{Rosa Félix\corref{cor1}%
  \corref{cor1}%
  }
   \ead{rosamfelix@tecnico.ulisboa.pt} 
    \author[CERIS]{Filipe Moura%
  %
  }
   \ead{fmoura@tecnico.ulisboa.pt} 
    \author[ITS]{Robin Lovelace%
  %
  }
   \ead{r.lovelace@uleeds.uk} 
      \affiliation[CERIS]{CERIS, Instituto Superior Técnico -
Universidade de Lisboa, Av Rovisco Pais 1, 1049-001 Lisboa, Portugal}
    \affiliation[ITS]{Institute for Transport Studies, University of
Leeds. 34-40 University Rd, Leeds LS2 9JT, UK}
    \cortext[cor1]{Corresponding author}

  \begin{abstract}
  A high proportion of car trips can be replaced by a combination of
  public transit and cycling for the first-and-last mile. This paper
  estimates the potential for cycling combined with public transit (PT)
  as a substitute for car trips in the Lisbon metropolitan area and
  assesses its socio-environmental impacts using open data and open
  source tools. A decision support tool that facilitates the design and
  development of a metropolitan cycling network was developed
  (\emph{biclaR}). The social and environmental impacts were assessed
  using the \emph{HEAT for Cycling} and the \emph{HEAT as a Service}
  tools. The impacts of shifting car trips to PT were also estimated and
  monetized. The results indicate that 10\% of trips could be made by
  bicycle + PT combination. Shifting to cycling for the first-and-last
  mile stages can reduce annual CO\textsubscript{2}eq emissions from
  3,000 to 7,500 tons/day, while for the PT leg, the transfer from car
  avoids of up to 20,500 tons of CO\textsubscript{2}eq emissions per
  year. The estimated socio-environmental benefits are of €125 million
  to €325 million over 10 years. This evidence can support policymakers
  to prioritize interventions that reduce the reliance on private motor
  vehicles.\\
  \end{abstract}
    \begin{keyword}
    Active transport \sep Intermodality \sep First and last
mile \sep Health economic assessment \sep Environmental impacts \sep 
    Open data and methods
  \end{keyword}

@@

To achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively) (Presidência do
Conselho de Ministros, 2019), the Lisbon's Metropolitan Department of
Transport introduced \emph{biclaR}\footnote{See
  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
decision support tool that facilitates the planning, design, and
development of a metropolitan cycling network (Félix et al., 2022).

\emph{biclaR} builds on the Propensity to Cycle Tool\footnote{See
  \href{https://www.pct.bike/}{pct.bike}.} (PCT), a web application and
research project funded by the UK's Department for Transport in 2015
which launched nationally in 2017 as part of the government's Cycling
and Walking Investment Strategy. The PCT initially used only
origin-destination data for commuting trips as the basis of estimates of
cycling potential at zone, route and route network levels (Lovelace et
al., 2017). The PCT has been extended to include cycling potential for
travel to school in England (Goodman et al., 2019) and other trip types
in other countries.\footnote{See \href{https://www.npt.scot}{npt.scot}
  and \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
  Scotland and Ireland that include estimates of cycling for other
  purposes.} However, to the best of our knowledge, this is the first
time that the method has been integrated with public transport data
using multi-modal routing to estimate the potential and benefits of
multi-stage cycling and PT trips.

@@

We used a method for disaggregating the origins and destinations of
trips between the centroids of two districts (same as ``parish'') to
ensure that a district is not solely characterized by a single point of
origin and destination for its trips. Aggregating all trips into
centroids renders the exercise less realistic, as it excludes a
significant portion of short-distance trips, a prevalent characteristic
of active mode travel (Lovelace et al., 2022b). The OD Jittering method
breaks down a single point (i.e., the centroid of an area) into multiple
random points on the existing and neighboring road network, using
OpenStreetMap as a reference. This method then distributes the volume of
trips within the district among the randomly generated
origin-destination pairs.

@@

\caption{Representation of OD pairs in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).}\label{fig:jitter}
\end{figure}

@@

The intermodality scenario considers trips combining PT and cycling for
the first and last legs. In a conservative approach, we have restricted
our analysis to the first and last legs with a combined length of up to
5 km (for instance: 1 km from origin to interface \emph{A} plus 4 km
from interface \emph{B} to destination) or up to 25 minutes on bike.
Furthermore, we have imposed restrictions on PT usage to include only
trips with no PT transfers, and up to 2 hours (120 min). Additionally,
we have only included PT modes that can easily accommodate bicycles,
such as trains, ferries, trams, and inter-municipal bus lines equipped
with bike racks (Figure \ref{fig:map1}).

@@

\caption{Bike routes with highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).}\label{fig:map2}
\end{figure}

@@

For the \emph{cycling legs of the journey} (first and last legs),
socio-environmental impacts were estimated, using the HEAT for Cycling
tool v5.2 (Kahlmeier et al., 2017) from the World Health Organization,
and the
\href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
package}\footnote{\texttt{HEATaaS} is under development. For more
  information contact
  \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.}. The
use of this package made it possible to run multiple scenarios with few
changes in input values, making the interaction with HEAT more reliable
when reproducing runs.\\
The HEAT tool provided estimates on the shifting from car to cycling for
a short term time horizon (i.e., one year) and the long term (i.e., ten
years). It estimates the differences between two considered scenarios.
In this case: one baseline scenario, with data from the mobility survey,
and one cycling potential scenario in which targets of 4\% and 10\% of
cycling levels were achieved, transferred from car trips. We considered
two dimensions: \emph{social} --- including the physical activity, air
pollution exposure, and road casualties; and \emph{environmental} ---
including CO\textsubscript{2}eq emissions and other pollutants.

@@

Table \ref{tab:summary1} presents the LMA total daily trips that can be
made with cycling + TP combination (with the aforementioned
restrictions), the trips in the baseline scenario and corresponding new
trips to achieve the national strategy targets (4\% and 10\%), for
different route profiles. For the cycling legs of the journey (first and
last legs), the environmental avoided emissions and monetized
socio-environment (SE) benefits are presented in \ref{tab:summary1b},
resulting from replacing car trips with cycling.

@@

Table \ref{tab:summary21} shows the potential trips by PT mode to
replace the second leg of the journey, in combination with cycling.
Train offers the greatest potential for substitution (88\%). When
comparing the existing PT interfaces (Figure \ref{fig:map1}) with the
bike routes with highest potential to serve as first and last legs
(Figure \ref{fig:map2}) it becomes clear that the Train interfaces are
the ones that have the highest potential to attract car-to-PT
substituting trips, if their accessibility by bicycle is improved to be
safe.

@@

The sum of CO\textsubscript{2}eq avoided emissions from the potential
car trips shifted to bike (first-and-last legs) in combination with PT
(second leg) in the LMA is presented on Table \ref{tab:summaryall}, for
both national cycling strategy targets and routing profiles, and the
socio-environmental benefits monetized in €, for a 1-year and 10-year
time periods.

@@

The environmental impacts represent less than 2\% of the
socio-environmental benefits (in value) from replacing car trips to
bicycle in first-and-last legs. For the PT segment, we did not estimate
the social impacts from substituting car trips. One of the main
socio-environmetal benefits, valued after monetization, comes from the
increase in physical activity (Félix et al., 2023). Although are also
social benefits form shifting car trips to PT, its health benefits would
not be as high as shifting to cycling. The literature shows that the
Metabolic Equivalent Tasks (MET) for ``riding in a bus or a train'' is
1.3 plus the ``walking for transportation'' as 3.5, while ``driving a
car'' is 2.5 (Ainsworth et al., 2011). The difference between these
activities is not very obvious when compared to cycling, whose
equivalent MET is between 6.8 to 7.5. Nevertheless, future works should
also encompass the estimation of the social impacts for the PT leg of
the journey, shifting from car.

@@

Our findings suggest that cycling and PT \emph{in combination} could
viably replace 10\% of current LMA trips, with an additional 6\% of PT
journeys prone to further substitution.

@@

By making the research process publicly accessible in a code repository,
it enables the replication of similar estimates for socio-environmental
impacts, resulting from a modal shift from car to bicycle in combination
with PT, in other metropolitan areas.

\hypertarget{acknowledgements.}{%
\subsubsection*{Acknowledgements.}\label{acknowledgements.}}
\addcontentsline{toc}{subsubsection}{Acknowledgements.}

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
Félix, R., Lovelace, R., \& Moura, F. (2022). \emph{{biclaR - Ferramenta
de apoio ao planeamento da rede ciclável na área metropolitana de
Lisboa}}. {CERIS - Instituto Superior Técnico and Transportes
Metropolitanos de Lisboa}. \url{https://biclar.tmlmobilidade.pt}

>>>>

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={XXX: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}

@@

  \title{XXX: Estimating the socio-environmental impacts of car
substitution by bicycle and public transit using open tools}
    \author[CERIS]{Rosa Félix\corref{cor1}%
  \corref{cor1}%
  }
   \ead{rosamfelix@tecnico.ulisboa.pt} 
    \author[CERIS]{Filipe Moura%
  %
  }
   \ead{fmoura@tecnico.ulisboa.pt} 
    \author[ITS]{Robin Lovelace%
  %
  }
   \ead{r.lovelace@uleeds.uk} 
      \affiliation[CERIS]{CERIS, Instituto Superior Técnico -
Universidade de Lisboa, Av Rovisco Pais 1, 1049-001 Lisboa, Portugal}
    \affiliation[ITS]{Institute for Transport Studies, University of
Leeds. 34-40 University Rd, Leeds LS2 9JT, UK}
    \cortext[cor1]{Corresponding author}

  \begin{abstract}
  A high proportion of car trips can be replaced by a combination of
  public transit and cycling for the first-and-last mile. This paper
  estimates the potential for cycling combined with public transit (PT)
  as a substitute for car trips in the Lisbon metropolitan area and
  assesses its socio-environmental impacts using open data and open
  source tools. A decision support tool that facilitates the design and
  development of a metropolitan cycling network was developed
  (\emph{XXX}). The social and environmental impacts were assessed using
  the \emph{HEAT for Cycling} and the \emph{HEAT as a Service} tools.
  The impacts of shifting car trips to PT were also estimated and
  monetized. The results indicate that 10\% of trips could be made by
  bicycle + PT combination. Shifting to cycling for the first-and-last
  mile stages can reduce annual CO\textsubscript{2}eq emissions from
  3,000 to 7,500 tons/day, while for the PT leg, the transfer from car
  avoids of up to 20,500 tons of CO\textsubscript{2}eq emissions per
  year. The estimated socio-environmental benefits are of €125 million
  to €325 million over 10 years. This evidence can support policymakers
  to prioritize interventions that reduce the reliance on private motor
  vehicles.\\
  \end{abstract}
    \begin{keyword}
    Active transport \sep Intermodality \sep First and last
mile \sep Health economic assessment \sep Environmental impacts \sep 
    Open data and methods
  \end{keyword}

@@

To achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively) (Presidência do
Conselho de Ministros, 2019), the Lisbon's Metropolitan Department of
Transport introduced \emph{XXX}\footnote{See https://xxx.xxx. \emph{For
  double blind peer-review process, the authors decided to truncate the
  name and references to this tool}.}, a decision support tool that
facilitates the planning, design, and development of a metropolitan
cycling network (ref X).

\emph{XXX} builds on the Propensity to Cycle Tool\footnote{See
  \href{https://www.pct.bike/}{pct.bike}.} (PCT), a web application and
research project funded by the UK's Department for Transport in 2015
which launched nationally in 2017 as part of the government's Cycling
and Walking Investment Strategy. The PCT initially used only
origin-destination data for commuting trips as the basis of estimates of
cycling potential at zone, route and route network levels (Lovelace et
al., 2017). The PCT has been extended to include cycling potential for
travel to school in England (Goodman et al., 2019) and other trip types
in other countries.\footnote{See \href{https://www.npt.scot}{npt.scot}
  and \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
  Scotland and Ireland that include estimates of cycling for other
  purposes.} However, to the best of our knowledge, this is the first
time that the method has been integrated with public transport data
using multi-modal routing to estimate the potential and benefits of
multi-stage cycling and PT trips.

@@

We used a method for disaggregating the origins and destinations of
trips between the centroids of two districts (same as ``parish'') to
ensure that a district is not solely characterized by a single point of
origin and destination for its trips. Aggregating all trips into
centroids renders the exercise less realistic, as it excludes a
significant portion of short-distance trips, a prevalent characteristic
of active travel (Lovelace et al., 2022b). The OD Jittering method
breaks down a single point (i.e., the centroid of an area) into multiple
random points on the existing and neighboring road network, using
OpenStreetMap as a reference. This method then distributes the volume of
trips within the district among the randomly generated
origin-destination pairs.

@@

\caption{Representation of desire lines in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).}\label{fig:jitter}
\end{figure}

@@

The intermodality scenario considers trips combining PT and cycling for
the first and last legs. In a conservative approach, we have restricted
our analysis to the first and last legs with a combined length of up to
5 km (for example: 1 km from origin to interface \emph{A} plus 4 km from
interface \emph{B} to destination) or up to 25 minutes on bike
travel-time. Furthermore, we have imposed restrictions on PT usage,
limiting it to trips without PT transfers, and within a duration of up
to 2 hours (120 minutes). Additionally, we have only included PT modes
that can easily accommodate bicycles, such as trains, ferries, trams,
and inter-municipal bus lines equipped with bike racks (Figure
\ref{fig:map1}). These restrictions can be eased in the future when
testing more developed policy interventions to enhance intermodality
between cycling and PT, considering both the vehicle and infrastructure
perspectives.

@@

\caption{Bike routes with the highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).}\label{fig:map2}
\end{figure}

@@

For the \emph{cycling legs of the journey} (first and last legs),
socio-environmental impacts were estimated, using the Health Economic
Assessment Tool (HEAT) for Cycling v5.2 (Kahlmeier et al., 2017), from
the World Health Organization, and the
\href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
package}\footnote{\texttt{HEATaaS} is under development. For more
  information contact
  \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.}. The
use of this package made it possible to run multiple scenarios with few
changes in input values, making the interaction with HEAT more reliable
when reproducing runs.\\
The HEAT tool provided estimates on the shifting from car to cycling for
a short term time horizon (i.e., one year) and the long term (i.e., ten
years). It estimates the differences between two considered scenarios.
In this case: one baseline scenario, with data from the mobility survey,
and one cycling potential scenario in which targets of 4\% and 10\% of
cycling levels were achieved, transferred from car trips. We considered
two dimensions: \emph{social} --- including the physical activity, air
pollution exposure, and road casualties; and \emph{environmental} ---
including CO\textsubscript{2}eq emissions and other pollutants.

@@

Table \ref{tab:summary1} presents the LMA total daily trips that can be
made with cycling + TP combination (with the aforementioned
restrictions), the trips in the baseline scenario and corresponding new
trips to achieve the national strategy targets (4\% and 10\%), for
different route profiles. For the cycling legs of the journey (first and
last legs), the environmental avoided emissions and monetized
socio-environment (SE) benefits are presented in Table
\ref{tab:summary1b}, resulting from replacing car trips with cycling.

@@

Table \ref{tab:summary21} shows the potential trips by PT mode to
replace the second leg of the journey, in combination with cycling.
Train offers the greatest potential for substitution (88\%). When
comparing the existing PT interfaces (Figure \ref{fig:map1}) with the
bike routes with highest potential to serve as first and last legs
(Figure \ref{fig:map2}) it becomes clear that the Train interfaces are
the ones that have the highest potential to attract car-to-PT
substituting trips, if their accessibility by bicycle is improved to
become safer.

@@

The sum of CO\textsubscript{2}eq avoided emissions from the potential
car trips shifted to bike (first-and-last legs) in combination with PT
(second leg) in the LMA is presented in Table \ref{tab:summaryall}, for
both national cycling strategy targets and routing profiles, and the
socio-environmental benefits monetized in €, for a 1-year and 10-year
time periods.

@@

The environmental impacts represent less than 2\% of the
socio-environmental benefits (in value) from replacing car trips to
bicycle in first-and-last legs. For the PT segment, we did not estimate
the social impacts from substituting car trips. One of the main
socio-environmetal benefits, valued after monetization, comes from the
increase in physical activity (Félix et al., 2023). Although there are
also social benefits form shifting car trips to PT, its health benefits
would not be as high as shifting to cycling. The literature shows that
the Metabolic Equivalent Tasks (MET) for ``riding in a bus or a train''
is 1.3 plus the ``walking for transportation'' as 3.5, while ``driving a
car'' is 2.5 (Ainsworth et al., 2011). The difference between these
activities - shifting from car to PT - is not very obvious when compared
to shifting from car to cycling, whose MET is about 6.8. Nevertheless,
future works should also encompass the estimation of the social impacts
for the PT leg of the journey, shifting from car.

@@

Our findings suggest that cycling \emph{in combination} with PT could
potentially replace 10\% of current LMA trips, with an additional 6\% of
PT journeys prone to further substitution.

@@

By making the research process publicly accessible in a code repository
(see \protect\hyperlink{research-data}{Research Data}), it enables the
replication of similar estimates for socio-environmental impacts,
resulting from a modal shift from car to bicycle in combination with PT,
in other metropolitan areas.

\hypertarget{acknowledgements}{%
\subsection*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{subsection}{Acknowledgements}

@@

\hypertarget{research-data}{%
\subsection*{Research data}\label{research-data}}
\addcontentsline{toc}{subsection}{Research data}

The data and the code to produce XXX and this paper are available at:
https://github.com/{[}\emph{blind}{]}.

@@
